### 2015-12-05 paper/1-introduction.tex
More rewrites & fixes.

%? use paragraphs for layout instead of (dense) flowing text ?
%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
\subsection{DEVS}
%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms. Of interest in this paper are the 2 key formalisms: Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation. This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by connected ports.

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a PDEVS simulation can lower overall runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

\subsection{Motivation}
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
Adevs \cite{adevs} offers a very fast conservative synchronized shared memory DEVS simulator, but no optimistic synchronized variant. 
The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code and without losing access to powerful low-level threading primitives. % may need another pass on this one :(

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. Dxex kernels can operate on IEEE754 floating point time units or integer time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
An integer representation significantly reduces the possible range of the virtual time, but avoids approximation errors. Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established. This task is non-trivial for floating point approximations.\\
Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernels ($2^{24}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This allows a deterministic ordering of concurrent events. % execution not so much as tracing

>>>>

%? use paragraphs for layout instead of (dense) flowing text ?
%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
\subsection{DEVS}
%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms. Of interest in this paper are the 2 key formalisms: Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation. This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by ports.

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a PDEVS simulation can lower overall runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not raise the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

\subsection{Motivation}
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
Adevs \cite{adevs} offers a very fast conservative synchronized shared memory PDEVS simulator, but no optimistic synchronized variant. 
The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code and without losing access to powerful low-level threading primitives. % may need another pass on this one :(

\subsection{Solution} 
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages. In practice, dxex defaults to word sized unsigned integer.
Although this significantly reduces the possible range of virtual time, it avoids approximation errors. Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established. This task is non-trivial for floating point approximations.\\
Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernels ($2^{48}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This allows a deterministic ordering of concurrent events. % execution not so much as tracing

### 2015-12-05 paper/3-features.tex
More rewrites & fixes.

%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
%TODO (maybe: note that models are not compatible)
\subsection{Different Synchronization protocols}
%TODO explain rationale: why different synchronization protocols
% Multiple synchronization protocols exist, with each their own strengths and weaknesses.
% weak scenarios for C, O and both ?
%TODO !!!!!!! This needs better explaining, perhaps a central graph to reference to ?
\subsubsection{Conservative}
A conservative kernel will determine which kernels it is influenced by. This information is constructed from the incoming connections on all hosted models. The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event. This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel. \\
The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.\\
%This is one of the operations that can benefit from using a relaxed memory ordering. % Using it for read atm, check before finalizing and do we need to mention this ?. %writing this here seems inappropriate
%TODO indicate in which cases conservative is good: perfect/good lookahead calculation (might be very hard) and what may be a problem: runtime behavior compared to static analysis makes the simulator too conservative. (e.g. a model might send a message, but never does will slow down the simulation)

>>>>

%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
%TODO (maybe: note that models are not compatible)
\subsection{Different Synchronization protocols}
%TODO explain rationale: why different synchronization protocols
Synchronization is required for PDEVS to prevent or recover from causality violations if the simulation is executed in parallel. Preventing causality violations (conservative) typically requires domain-specific information to compensate for performance loss, while recovering from causality errors (optimistic) requires a state to recover to.

\subsubsection{Conservative}
A conservative kernel will determine which kernels it is influenced by. This information is constructed from the incoming connections on all hosted models. The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event. This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel. \\
The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.\\
%This is one of the operations that can benefit from using a relaxed memory ordering. % Using it for read atm, check before finalizing and do we need to mention this ?. %writing this here seems inappropriate

### 2015-12-05 paper/4-performance.tex
More rewrites & fixes.

%TODO say that you will now prove your claims about the aproach having nearly no impact
%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
\subsection{Sequential Simulation}
%TODO we start with sequential simulation results, because this is still base case
%TODO compare CPU time with adevs for some benchmarks
%TODO and explain why there is such a difference
%TODO (spend some time here, as this is also important)
\subsubsection{Preamble}
The benchmark models are identical in functionality (between adevs/dxex versions). Dxex uses integer timestamps, adevs uses floating point. Tests show there is little difference performance wise (both types fit in a machine word)\\
Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel). Analysis of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository. \\
Note that a fixed timeadvance (e.g. devstone) is set at 100.
\subsubsection{Devstone}
The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this whereas adevs needs to walk the coupled model's structure to pass events.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
	\label{fig1.eps}
\end{figure}

@@

\subsubsection{Priority network model}
%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers. \\ This defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph. This model therefore highlights the basic strengths/weaknesses of both synchronization protocols. Receiving models are allocated on another kernel than the server, and have a internal transition so that they will not wait for the incoming messages. 
\input pqueue.tex \\
A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create. The kernel holding the server will never revert since it is a source in the dependency graph. Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold. The overhead in optimistic is entirely due to the factor m, which will quickly dominate in increasing buffers of received messages.
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig5.eps}
	\label{fig5.eps}
\end{figure}
Interestingly, the parameter p does not clearly benefit either synchronization protocol. While this removes any possibility for a lookahead, conservative can quickly bridge the timespan between fixed messages. Lightweight states prevent optimistic from performance loss due to reverts, only the overhead in storing messages becomes dominant.

@@

\paragraph*{Priority Network}
The priority network model is detailed in section 4.1.\\
Conservative's high memory usage is largely due to the garbage collection implementation. Kernel 0 is the initiator of the lbts calculation but since it holds only models independent of any other (server), it will finish simulation very fast leaving the other kernel bereft of an updated lbts. The kernel holding the server will after simulation wait on the receiving kernel until it can prove all send messages can be deallocated.
\begin{table}[lhtb]
	\centering
	\caption{Priority model n 128, m 16, p 10,  t2e8, unit MiB, 2 kernels}
	\label{pmod_mem}
	\begin{tabular}{| l | l | l |}
		\hline
		DX &DXCon &DXOpt\\ \hline
		35 & 450 & 651\\ \hline
	\end{tabular}
\end{table}

>>>>

%TODO say that you will now prove your claims about the aproach having nearly no impact
%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
\subsection{Sequential Simulation}
%TODO we start with sequential simulation results, because this is still base case
%TODO compare CPU time with adevs for some benchmarks
%TODO and explain why there is such a difference
%TODO (spend some time here, as this is also important)
\subsubsection{Preamble}
The benchmark models are identical in functionality (between adevs/dxex versions). Dxex uses integer timestamps, adevs uses floating point. Tests show there is little difference performance wise (both types fit in a machine word)\\
Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel), G++ 5.2. Interpretation of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository. For the runtime plots perf is used in combination with R.\\
Note that a fixed timeadvance (e.g. devstone) is set at 100.
\subsubsection{Devstone}
The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this whereas adevs needs to walk the coupled model's structure to pass events.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
	\label{fig1.eps}
\end{figure}

@@

\subsubsection{Priority network model}
%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers. \\ This defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph. This model therefore highlights the basic strengths/weaknesses of both synchronization protocols. Receiving models are allocated on another kernel than the server, and have a internal transition so that they will not wait for the incoming messages. 
\input pqueue.tex \\
A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create. The kernel holding the server will never revert since it is a source in the dependency graph. Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold. The overhead in optimistic is entirely due to the factor m, which will quickly dominate in increasing buffers of received messages.
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig5.eps}
	\label{fig5.eps}
\end{figure}
Interestingly, the parameter p does not clearly benefit either synchronization protocol. While this removes any possibility for a lookahead, conservative can quickly bridge the timespan between fixed messages since there is no cyclic dependency. Lightweight states prevent performance loss due to reverts in optimistic, only the overhead in event handling in optimistic eventually becomes the deciding factor.

@@

\paragraph*{Priority Network}
The priority network model is detailed in section 4.1.\\
Conservative's high memory usage is largely due to the garbage collection implementation. Kernel 0 is the initiator of the lbts calculation but since it holds only models independent of any other (server), it will finish simulation very fast leaving the other kernel bereft of an updated lbts. The kernel holding the server will after simulation wait on the receiving kernel until it can prove all sent messages can be deallocated.
\begin{table}[lhtb]
	\centering
	\caption{Priority model n 128, m 16, p 10,  t2e8, unit MiB, 2 kernels}
	\label{pmod_mem}
	\begin{tabular}{| l | l | l |}
		\hline
		DX &DXCon &DXOpt\\ \hline
		35 & 450 & 651\\ \hline
	\end{tabular}
\end{table}
